---
title: "Bayesian Analysis of Hi-C Interactions with HiCPotts"
author: "Itunu Godwin Osuntoki, Nicolae Radu Zabet"
date: "`r Sys.Date()`"
output:   
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Bayesian Analysis of Hi-C Interactions with HiCPotts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

High-throughput chromosome conformation capture (Hi-C) technologies have
revolutionized our understanding of 3D genome organization by mapping
interactions between genomic loci. However, Hi-C data are inherently noisy and
affected by experimental biases such as GC content, transposable elements, and
DNA accessibility, which complicate the identification of biologically
significant interactions.

The HiCPotts package provides a comprehensive framework for Bayesian analysis of
Hi-C interaction data using a Hidden Markov Random Field (HMRF) model. Hi-C is a
high-throughput sequencing technique that captures chromatin interactions across
the genome, revealing spatial organization. This package models these
interactions using a mixture of distributions (Poisson, Negative Binomial,
Zero-Inflated Poisson, or Zero-Inflated Negative Binomial) while accounting for
covariates, genomic distance, GC content, accessibility, and transposable
element (TE) counts. The HMRF framework incorporates spatial dependencies via a
Potts model, and the package employs Markov Chain Monte Carlo (MCMC) methods for
parameter estimation.

An HMRF models spatial dependencies in a lattice (e.g., a matrix of Hi-C
interactions) by assigning each site to one of several latent states (mixture
components). The Potts model governs spatial interactions, encouraging
neighboring sites to share the same state, controlled by an interaction
parameter ($\gamma$).

The package uses MCMC to estimate model parameters (regression coefficients
($\beta$), zero-inflation parameter ($\theta$), dispersion parameter for
Binomial distributions, and ($\gamma$)) by sampling from their posterior
distributions. Covariates like genomic distance and GC content adjust for biases
in interaction counts. Each Hi-C interaction is modeled as belonging to one of
three mixture components following a specified distribution.

The package performance was optimized through C++ implementations using Rcpp and
RcppArmadillo. The package also supports parallel processing and flexible
distribution choices, making it suitable for large-scale genomic analyses.

# Motivation

Most existing computational methods fail to adequately model the spatial
dependencies and overdispersion in Hi-C contact matrices, limiting their ability
to distinguish true signals from other components such as noise. The HiCPotts
package addresses these challenges by providing a novel Bayesian framework to
detect enriched interactions while accounting for experimental biases in Hi-C
data. Its integration into Bioconductor is motivated by its robust statistical
approach, computational efficiency, and ability to provide insights into bias
sources, making it a valuable tool for researchers studying chromatin
architecture in diverse biological contexts.

HiCPotts package also extends the existing knowledge of classifying interacting
loci into two components to three components (true signal, false signal, and
noise) and its focus on bias correction (DNA accessibility, transposable
elements) enhance its utility for integrative genomic studies.

# Comparison with Existing Packages
There are several Bioconductor packages that addresses Hi-C data analysis, each
with distinct functionalities and scopes. We highlight the uniqueness of the
HiCPotts package with other existing packages:

a. diffHic: The diffHic package focuses on detecting differential interactions
between biological conditions using the edgeR framework for statistical
modeling. diffHic provides methods for read pair alignment, binning, filtering,
and normalization of biases (e.g., trended or CNV-driven). While diffHic excels
at differential analysis, it does not explicitly model spatial dependencies in
Hi-C data or account for overdispersion as HiCPotts does. HiCPotts is better
suited for identifying enriched interactions within a Hi-C experiment and
exploring bias sources, whereas diffHic is ideal for comparative studies across
conditions.


b. HiCcompare and multiHiCcompare: HiCcompare offers joint normalization and
difference detection for multiple Hi-C datasets, operating on sparse chromatin
interaction matrices. multiHiCcompare extends this to handle multiple groups and
replicates using cyclic loess normalization and a general linear model (GLM)
based on edgeR. Both packages emphasize comparative analysis and normalization
but do not focus on detecting enriched interactions within a dataset or modeling
spatial dependencies. HiCPotts’s Bayesian approach and bias correction make it
complementary, as it prioritizes significant interaction detection and bias
insight over differential analysis.

c. HiCDCPlus: HiCDCPlus enables significant interaction calling and differential
analysis for Hi-C and HiChIP data using a negative binomial generalized linear
model. It includes tools for topologically associating domain (TAD) and A/B
compartment calling, integrating with visualization tools. Like HiCPotts, it
calls significant interactions, but HiCPotts’s HMRF-based model and ABC
approach
provide superior handling of spatial dependencies and computational tractability
for chromosome-wide analysis. HiCDCPlus requires GC content information
(computable internally), while HiCPotts additionally corrects for transposable
elements and DNA accessibility biases.

d. scHiCcompare: Designed for single-cell Hi-C data, scHiCcompare supports
imputation, normalization, and differential interaction analysis across
single-cell datasets. Its focus on single-cell data makes it distinct from
HiCPotts, which targets bulk Hi-C data. HiCPotts’s ability to model
overdispersion and spatial dependencies is not replicated in scHiCcompare, which
prioritizes single-cell-specific challenges.


Other Tools: Packages like HiCdat and HiCExperiment provide preprocessing,
visualization, or data manipulation for Hi-C data but lack the statistical rigor
of HiCPotts for interaction detection. HiCdat offers a graphical interface for
preprocessing and integrative analysis with other omics data, while
HiCExperiment provides data structures for 3C-related experiments. Neither
focuses on enriched interaction detection or bias correction like HiCPotts.

# Features of HiCPotts

HiCPotts allows researchers to identify significant intra-chromosomal
interactions in Hi-C data while correcting for experimental biases. Its key
features include:

1.  Use of Zero-Inflated distributions to handle overdispersion.

2.  An HMRF-based Bayesian framework with the Potts model for spatial
    dependency.

3.  ABC handles the Potts model’s normalizing constant efficiently
    constant.

4.  Bias correction for GC content, transposable elements, and DNA
    accessibility.

5.  Unlike diffHic, HiCcompare, and multiHiCcompare, which focus on differential
    analysis, or HiCDCPlus, which balances interaction calling and differential
    analysis, HiCPotts prioritizes enriched interaction detection. Its bias
    correction and spatial modeling make it a powerful complement to existing
    tools, enhancing Bioconductor’s suite for 3D genome analysis. HiCPotts can
    be used for initial interaction detection, followed by diffHic or HiCcompare
    for differential studies, creating a comprehensive Hi-C analysis workflow.


# Installation

HiCPotts depends on several CRAN and Bioconductor packages.

Install them as follows:

Install CRAN dependencies


```{r setup, include=FALSE}
library(BiocStyle)
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, collapse = TRUE)
```

```{r install-cran, echo=TRUE, eval=FALSE}
install.packages(c("Rcpp", "RcppArmadillo", "parallel"))
```

```{r install-bioc, echo=TRUE, eval=FALSE}
# Install Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install(c(
    "rhdf5", "strawr", "rtracklayer",
    "GenomicRanges", "BSgenome", "Biostrings"
))
```

```{r load-package, echo=TRUE, eval=TRUE}
# Loading the package:
# install.packages(HiCPotts)
library(HiCPotts)
```

# Workflow Overview

The HiCPotts workflow involves five main steps:

1.  Data Loading: Use `get_data()` to read HDF5-backed Hi-C contact matrices
    (`.cool` and supported `.h5` schemas) and annotate bins with GC content,
    accessibility, and TE counts.

2.  Data Processing: Use `process_data()` to convert the data into (N \times N)
    matrices of interactions and covariates.

3.  MCMC Simulation: Use `run_chain_betas()` to run MCMC chains, estimating
    parameters and latent state assignments.

4.  Classification: Use `classify_hicpotts()` to pool post-burn-in latent-state
    frequencies and assign every interaction to noise, signal, or false signal.
    Parameter-plus-Potts probabilities are available separately through
    `compute_HMRFHiC_probabilities()` as a sensitivity check.

5.  Visualization: Use `plot_upper_prob_lower_count()` to create a dual-triangle
    Hi-C heatmap that overlays posterior probabilities (upper triangle, blue
    gradient) and observed interaction counts (lower triangle, grey gradient)
    for intuitive result interpretation.

For an argument-by-argument guide to every exported function, including return
objects, internal processing stages and runnable examples, see
`vignette("HiCPotts_function_reference", package = "HiCPotts")`.

We’ll demonstrate this workflow using synthetic Hi-C data for simplicity.

## Step 1: Loading Hi-C Data

The `get_data()` function reads Hi-C contact matrices and annotates bins with
covariates. For this example, we simulate a small dataset instead of using real
Hi-C files, which require specific file formats and genome annotations.

Simulate a 10x10 Hi-C dataset

```{r simulate-data, include=FALSE, eval=TRUE}
set.seed(4921)

# Lattice size: an N x N grid of genomic bins.  Each cell of the grid
# represents an interaction between bin i and bin j.
N <- 10

# Define the start coordinate of each bin (10 bins, 2 kb wide →
# 20 kb region)
bins <- seq(1, 20000, by = 2000)

# Create every pairwise combination of bin i vs bin j
data <- expand.grid(start = bins, start.j. = bins)

# Bin end coordinates (each bin is 2 kb wide)
data$end.i. <- data$start + 1999

data$end <- data$start.j. + 1999

# Chromosome label (Drosophila chr2L here)
data$chrom <- "chr2L"

# ── Simulate the four bias covariates
# --- Simulate covariates and counts
# Simulated GC content, GC fraction (biological range 30–70%)
data$GC <- runif(nrow(data), 0.3, 0.7)

# Simulated accessibility, DNase-I accessibility score (0–1)
data$ACC <- runif(nrow(data), 0, 1)

# Simulated TE counts, count of overlapping transposable elements
data$TES <- rpois(nrow(data), 2)

# Poisson(λ = 5) approximates sparse Hi-C counts at this resolution
# Simulated interaction counts
data$interactions <- rpois(nrow(data), lambda = 5)

# Coerce to a plain data.frame (expand.grid returns a class with extra
# attrs)
data <- as.data.frame(data)

# Inspect the first few rows – note the required column names:
# start, start.j., end.i., end, chrom, GC, ACC, TES, interactions
head(data)
```

In practice, the `get_data()` function is used to load real Hi-C data from a
`.cool` file or a supported `.h5` schema. Binary `.hic` files must first be
converted to `.cool` or read separately with `strawr::straw()`. For `.mcool`,
select and export the required resolution as `.cool` before calling
`get_data()`. These formats can also be represented by other Bioconductor
containers, but their returned objects differ.
HiCExperiment generates a HiCExperiment object that includes a contact matrix,
genomic regions, metadata (e.g., resolution, chromosome), and pairwise
interactions. In contrast, `get_data()` produces a data frame from the same file
types, incorporating optional calculations (if available) to identify sources of
experimental biases, such as GC content or DNA accessibility. Also the function
support loading profiles as bigwig or bedgraph and will be imported as GRanges.

```{r get_data, include=FALSE, eval=FALSE}
# Example (not run)
# For actual Hi-C / Micro-C files use get_data() with these arguments:
#
#   file_path      – path to a .cool or supported .h5 matrix
#   chr            – chromosome to extract (e.g. "chr2L", "chr4")
#   start, end     – genomic window (bp, 1-based inclusive)
#   resolution     – target bin size in bp; native bins are merged/sliced
#   genome_package – BSgenome package used for GC computation
#   acc_wig        – optional bedGraph/wig of DNase-I (or ATAC) accessibility
#   chain_file     – optional liftOver chain if acc_wig is in another assembly
#   te_granges     – optional BED/GTF (or GRanges) of transposable elements

data <- get_data(
    file_path = "path/to/hic_file.h5",
    chr = "chr2L",
    start = 1,
    end = 20000,
    resolution = 2000,
    genome_package = "BSgenome.Dmelanogaster.UCSC.dm6",
    acc_wig = "path/to/dnase.wig",
    chain_file = "path/to/chain.file",
    te_granges = "path/to/te.gtf"
)
```

## Step 2: Processing Data

The `process_data()` function converts the data frame generated through the
`prepare_data()` function into a list of (N $\times$ N) matrices for
interactions (`y`) and covariates (`x_vars`), optionally scaling interaction
counts. Also, if a HiCExperiment object is already available and the sources of
biases are known and organized in a separate data frame, these can be combined
into a single data frame using base R functions and the `prepare_data()`
function to convert into a list for analysis.

```{r process_data, include=FALSE, eval=TRUE}
# ── Reshape the tidy interaction table into matrices
# for MCMC
# ────────────────
# process_data() returns a list with two elements: $y – an N x N
# matrix of observed interaction counts $x_vars – a list of N x
# N matrices, one per covariate (distance, GC, TES, ACC) Arguments: data
# – data.frame with the 9 required columns (see above) N
# – lattice dimension (here 10, matching 10 bins) scale_max
# – upper bound used when rescaling counts; only consulted when
# standardization_y = TRUE standardization_y – keep FALSE for
# count models. The Poisson/NB/ZIP/ZINB likelihoods require raw integer
# counts; min-max rescaling (TRUE) breaks the count assumption and is
# retained only for backwards compatibility.

processed <- process_data(data, N = N, standardization_y = FALSE)

# Pull the two pieces out of the returned list
x_vars <- processed[["x_vars"]]  # Single matrix for this example
y <- processed[["y"]]

# Inspect x_vars: should be a list of distance, GC, TES, ACC matrices
str(x_vars)  # Lists of distance, GC, TEs, ACC matrices
```

This produces matrices for distance, GC, TEs, ACC, and interaction counts, ready
for MCMC.

## Step 3: Running MCMC Simulations

The `run_chain_betas()` function runs MCMC simulations to estimate parameters
(($\beta$), ($\gamma$), ($\theta$ (for Zero-Inflated distributions)), size(for
Negative binomial distributions)) and latent state assignments ((z)). In the
example below, we use the Zero-Inflated Negative Binomial (ZINB) distribution to
model the simulated sparse Hi-C data.

```{r setup_MCMC, include=FALSE, eval=TRUE}
# ── Define MCMC hyper-parameters and starting values
# ─────────────────────────

# Potts spatial-coupling parameter (γ). Values closer to 1 encourage
# neighbouring bins on the lattice to share the same latent state.
gamma_start <- 0.3

# Total number of MCMC iterations. Burn-in defaults to iterations / 2.  Use
# a small value here for vignette speed; 5 000+ is typical for real data.
iterations <- 20

# Initial value for the zero-inflation parameter (θ).  Required when
# dist = 'ZIP' or 'ZINB'; ignored otherwise.
theta_start <- c(0.5)

# size_start (commented out) – only needed for 'NB' or 'ZINB':
# one initial over-dispersion value per mixture component.  size_start <-
# c(1, 1, 1)

# Choice of count distribution for interactions. Options: 'Poisson', 'NB',
# 'ZIP', 'ZINB' ZIP is a reasonable default for sparse Hi-C with many
# structural zeros.
dist <- "ZIP"
```

```{r run_MCMC, include=FALSE, eval=TRUE}
# --- Run the Metropolis-Hastings MCMC sampler
# run_chain_betas() samples from the joint posterior of the regression
# coefficients for all three mixture components simultaneously.

results <- run_chain_betas(
    N = N,
    gamma_start = gamma_start,
    iterations = iterations,
    x_vars = x_vars,
    y = y,
    theta_start = theta_start,
    use_data_priors = TRUE,
    # user_fixed_priors = NULL,
    # epsilon = NULL,
    distance_metric = "manhattan",
    dist = dist,
    # size_start = size_start,
    mc_cores = 1
)
```

```{r MCMC_results, include=FALSE, eval=TRUE}
# ── Extract the individual components of the MCMC
# output
# ─────────────────────
# 'results' is a named list. Each element corresponds to a different set of
# posterior samples.

# Regression coefficient chains – one matrix per mixture
# component (rows = iterations, columns = covariates incl. intercept)

chains <- results

# Posterior samples of the Potts spatial-coupling parameter γ
gamma <- results[[1]][["gamma"]]

# Posterior samples of the zero-inflation parameter θ (ZIP / ZINB
# only)
theta <- results[[1]][["theta"]]

# Over-dispersion size parameter – only present for NB / ZINB
# size <- results[['size]]
```

The output includes chains for regression parameters (`chains`), the Potts
interaction parameter (`gamma`), zero-inflation parameter (`theta`), and
dispersion parameters (`size`).

## Step 4: Official Three-Component Classification

The official classification comes from the latent states sampled by the full
emission-plus-Potts model. `classify_hicpotts()` pools post-burn-in membership
frequencies and returns exactly three biological labels: noise, signal and
false signal. Confidence diagnostics do not create another class.

```{r compute_result, include=FALSE, eval=TRUE}
probs <- classify_hicpotts(chains, data = data, min_draws = 5L)

# Matrices of probabilities for the first component, which translate
# to assigning interactions to biological states (e.g., active vs.
# inactive chromatin).
probs$prob1

# Optional parameter-plus-Potts check; this never replaces `probs`
# or its labels.
parameter_probs <- compute_HMRFHiC_probabilities(
    data = data, chain_betas = chains, iterations = iterations,
    N = N, dist = "ZIP", relabel = TRUE
)
head(parameter_probs[, c("prob1", "prob2", "prob3")])
```

## Step 5: Parameter Summaries and Diagnostics

HiCPotts reports posterior estimates together with intervals, effective
sample sizes and split-chain R-hat. `diagnose_hicpotts_fit()` records each
criterion and its threshold in `reliability_flags`.
`summarise_hicpotts_parameters()` applies the reporting criteria described
in its help page; classification has a separate cell-level summary.

Cross-chain convergence assessment uses independently seeded chains.
`process_data()` wraps its
count matrix in a length-1 list, so pass `y[[1]]` to fit one dataset:

```{r diagnose_fit, echo=TRUE, eval=TRUE}
multi <- run_chain_betas(
    N = N, gamma_start = gamma_start, iterations = iterations,
    x_vars = x_vars, y = y[[1]], theta_start = theta_start,
    use_data_priors = TRUE, dist = dist,
    n_chains = 2, seeds = 1:2, mc_cores = 1
)

diagnostics <- diagnose_hicpotts_fit(multi)
diagnostics$reliability_flags[, c("criterion", "passed")]
```

This 20-iteration example demonstrates the interface. Its ESS can be
compared directly with the default threshold of 200; production analyses
use a chain length chosen for the desired Monte Carlo precision. See
`?diagnose_hicpotts_fit` for the definition of every recorded criterion.
The following call displays estimates and the row-level diagnostic result:

```{r summarise_parameters, echo=TRUE, eval=TRUE}
parameter_summary <- summarise_hicpotts_parameters(multi,
    require_reliable = FALSE
)
head(parameter_summary[, c(
    "parameter", "estimate", "CI_lower", "CI_upper", "resolved"
)])
```

For threshold-enforced reporting, use `require_reliable = TRUE`.
The `resolved` column records whether each row meets its ESS/R-hat criteria;
the attached metadata records the other diagnostic results.

Two further checks are worth running before interpreting results.
`allocation_diagnostics()` reports Monte Carlo error and effective sample
size for the membership probabilities themselves, cell by cell, so a
probability sitting near a decision boundary can be told apart from one
that is merely under-sampled:

```{r allocation_diagnostics, echo=TRUE, eval=TRUE}
alloc <- allocation_diagnostics(multi)
names(alloc)

# Effective sample size of the worst-resolved lattice cell.
alloc$worst_cell_ess
```

`posterior_predictive_hicpotts()` answers the complementary question of
whether the fitted model reproduces the observed count distribution at
all. Each row compares an observed summary with the range obtained from
replicated datasets simulated from the posterior:

```{r posterior_predictive, echo=TRUE, eval=TRUE}
pp <- posterior_predictive_hicpotts(
    multi,
    x_vars = x_vars, y = y[[1]], dist = dist, n_rep = 20L
)
pp$summary
```

## Step 6: Visualizing Results

The `plot_upper_prob_lower_count()` function produces a symmetric Hi-C heatmap
where the two triangles convey complementary information:

- **Upper triangle** — posterior probability of belonging to the
    elevated-signal
    component (`prob2` by default), rendered in a blue gradient. Dark blue
    indicates a high probability of enriched interaction.
- **Lower triangle** — observed interaction counts (or their `log1p`
    transform),
    rendered in a grey gradient. Darker shading corresponds to higher counts.

Plotting both layers in a single figure makes it straightforward to assess where
the model assigns high signal probabilities and whether those assignments are
supported by the raw counts.

The function accepts the data frame returned by `classify_hicpotts()` directly.
Supply the column names for the two bin coordinates (`bin1_col` and `bin2_col`),
the probability column (`prob_col`), and the count column (`count_col`). For the
synthetic dataset the bin start coordinates are stored in `start` (bin *i*) and
`start.j.` (bin *j*).

Set `symmetric_matrix = TRUE` for a Hi-C contact map that contains one complete
triangle or two mirrored copies of each contact. A full matrix is checked for
equal mirrored counts before plotting. Set `symmetric_matrix = FALSE` only when
the two orientations are intentionally different; in that mode the ordered
upper and lower cells are preserved without cross-triangle aggregation. The
synthetic example below samples cells independently, so it uses the ordered
mode; the genuine Hi-C and Micro-C examples use the symmetric mode.

```{r plot_results, echo=TRUE, eval=TRUE}
p <- plot_upper_prob_lower_count(
    results       = probs,
    bin1_col      = "start",
    bin2_col      = "start.j.",
    prob_col      = "prob2",
    count_col     = "interactions",
    chr_label     = "2L",
    title         = "Significant interactions detected by HiCPotts",
    use_log_count = TRUE,
    symmetric_matrix = FALSE
)
print(p)
```


# Worked Examples: Hi-C and Micro-C Data

The synthetic example above is useful for illustrating the API but is not
biologically meaningful. This section walks through two fully realistic
analyses — one for **Hi-C** and one for **Micro-C** — using the exact same
four-step workflow. The Hi-C data files are shipped with the package under
`inst/extdata`; the Micro-C paths are illustrative and should be replaced
with your own files.

## Example 1: Hi-C Data

We analyse a 100 kb-resolution Hi-C contact map of *D. melanogaster* chr4
in BG3 cells, integrating DNase-I accessibility and transposable-element
annotations for bias correction.

### Step 1 — Locate input files

```{r hic_paths, echo=TRUE, eval=TRUE}
# Hi-C contact matrix (.cool, 100 kb, BG3 cells, chr4)
hic_file <- system.file("extdata",
    "BG3_WT_merged_hic_matrix_chr4_100Kb.cool",
    package = "HiCPotts"
)

# DNase-I accessibility bedGraph (dm3 genome coordinates)
dnase_file <- system.file("extdata",
    "DNaseI_BG3_gr_chr4.bedGraph",
    package = "HiCPotts"
)

# LiftOver chain — maps dm3 accessibility up to dm6 assembly
chain_file <- system.file("extdata",
    "dm3ToDm6_chr4_only.chain",
    package = "HiCPotts"
)

# Transposable-element annotations (GTF, dm6 chr4)
te_file <- system.file("extdata",
    "dm6_TEs_chr4.gtf",
    package = "HiCPotts"
)

stopifnot(
    file.exists(hic_file),
    file.exists(dnase_file),
    file.exists(chain_file),
    file.exists(te_file)
)
```

### Step 2 — Load Hi-C data and compute covariates

```{r hic_get_data, echo=TRUE, eval=TRUE}
# get_data() does four things in one call:
#   1. Reads the .cool file and rebins to the requested resolution
#   2. Calculates GC content from the BSgenome package
#   3. Lifts over the DNase-I signal (dm3 -> dm6) and aggregates per bin
#   4. Counts transposable elements overlapping each bin pair
hic_data <- get_data(
    file_path      = hic_file,
    chr            = "chr4",
    start          = 1,
    end            = 400000,
    resolution     = 200000,
    genome_package = "BSgenome.Dmelanogaster.UCSC.dm6",
    acc_wig        = dnase_file,
    chain_file     = chain_file,
    te_granges     = te_file
)

head(hic_data)
```

### Step 3 — Process data into matrices

```{r hic_process, echo=TRUE, eval=FALSE}
# (end - start) / resolution, rounded up for the edge bin.
N_hic <- 14

processed_hic <- process_data(
    hic_data,
    N                 = N_hic,
    standardization_y = FALSE # raw counts: required by the count likelihoods
)

x_vars_hic <- processed_hic[["x_vars"]]
y_hic <- processed_hic[["y"]]
```

### Step 4 — Run MCMC with a realistic sampler configuration

```{r hic_mcmc, echo=TRUE, eval=FALSE}
# ZINB is recommended for Hi-C because of structural zeros and
# over-dispersion typical of ligation-based count data.
set.seed(2025)

hic_results <- run_chain_betas(
    N                = N_hic,
    gamma_start      = 0.3,
    iterations       = 5000,
    x_vars           = x_vars_hic,
    y                = y_hic,
    theta_start      = 0.5,
    size_start       = c(1, 1, 1),
    use_data_priors  = TRUE,
    distance_metric  = "manhattan",
    dist             = "ZINB",
    mc_cores         = min(4L, parallel::detectCores() - 1L)
)

# run_chain_betas() returns a list with one element per input dataset.
# Unpack the first (and here only) chain set:
hic_chains <- hic_results[[1]][["chains"]]
hic_gamma <- hic_results[[1]][["gamma"]]
hic_theta <- hic_results[[1]][["theta"]]
hic_size <- hic_results[[1]][["size"]]
```

### Step 5 — Official posterior component classification

```{r hic_probs, echo=TRUE, eval=FALSE}
# Pass the full results list, not just the coefficient matrices.
hic_probs <- classify_hicpotts(hic_results, data = hic_data)

# hic_probs is a data.frame with prob1, prob2, prob3 columns that sum to 1
# per row. prob1 is the noise/baseline component; prob2 and prob3 are the
# elevated-signal regimes.
head(hic_probs[, c("start", "end", "interactions", "prob1", "prob2", "prob3")])

# Bin-pairs confidently classified as elevated signal (prob2 > 0.9).
# This value can be increased or reduced based on the user's
# justification.
genuine_signal <- hic_probs[
    hic_probs$prob2 > 0.9,
    c("start", "end", "interactions", "prob2")
]
head(genuine_signal)
```

### Step 6 — Visualize results

`plot_upper_prob_lower_count()` takes the `hic_probs` data frame and renders a
dual-triangle heatmap. The upper triangle shows `prob2` (the elevated-signal
component probability) and the lower triangle shows log-transformed interaction
counts, so the fit and the data can be compared at a glance.

```{r hic_plot, echo=TRUE, eval=FALSE}
# Dual-triangle heatmap: upper = prob2 (blue), lower = log1p(count) (grey)
hic_plot <- plot_upper_prob_lower_count(
    results       = hic_probs,
    bin1_col      = "start",
    bin2_col      = "start.j.",
    prob_col      = "prob2",
    count_col     = "interactions",
    chr_label     = "chr4",
    title         = "Hi-C significant interactions – BG3 chr4 (100 kb)",
    use_log_count = TRUE,
    symmetric_matrix = TRUE
)
print(hic_plot)
```

## Example 2: Micro-C Data

Micro-C uses micrococcal nuclease (MNase) instead of a restriction enzyme,
yielding nucleosome-resolution contact maps (~150 bp). The HiCPotts workflow
is **identical** to Hi-C; practical differences:

* **Input file**: typically a higher-resolution `.cool` or `.mcool`.
* **Resolution**: 1 kb – 5 kb rather than 10 – 100 kb.
* **Structural zeros**: even more abundant at fine resolution — ZINB strongly
    recommended.
* **Iterations**: larger lattices need longer chains to converge.

### Step 1 — Locate input files

```{r microc_paths, echo=TRUE, eval=FALSE}
# Replace these with the actual paths on your machine.
microc_file <- "path/to/MicroC_BG3_chr4_1kb.cool"
dnase_file <- "path/to/DNaseI_BG3_gr_chr4.bedGraph"
chain_file <- "path/to/dm3ToDm6_chr4_only.chain"
te_file <- "path/to/dm6_TEs_chr4.gtf"
```

### Step 2 — Load Micro-C data

```{r microc_get_data, echo=TRUE, eval=FALSE}
# Aggregate from the native 1 kb to 5 kb to keep the lattice manageable.
microc_data <- get_data(
    file_path      = microc_file,
    chr            = "chr4",
    start          = 1,
    end            = 1350000,
    resolution     = 5000,
    genome_package = "BSgenome.Dmelanogaster.UCSC.dm6",
    acc_wig        = dnase_file,
    chain_file     = chain_file,
    te_granges     = te_file
)

head(microc_data)
```

### Step 3 — Process into matrices

```{r microc_process, echo=TRUE, eval=FALSE}
# 1,350,000 / 5,000 = 270 bins -> a 270 x 270 lattice.
N_microc <- 270

processed_microc <- process_data(
    microc_data,
    N                 = N_microc,
    standardization_y = FALSE
)

x_vars_microc <- processed_microc[["x_vars"]]
y_microc <- processed_microc[["y"]]
```

### Step 4 — Run MCMC

```{r microc_mcmc, echo=TRUE, eval=FALSE}
set.seed(2025)

microc_results <- run_chain_betas(
    N                = N_microc,
    gamma_start      = 0.3,
    iterations       = 10000,
    x_vars           = x_vars_microc,
    y                = y_microc,
    theta_start      = 0.5,
    size_start       = c(1, 1, 1),
    use_data_priors  = TRUE,
    distance_metric  = "manhattan",
    dist             = "ZINB",
    mc_cores         = min(4L, parallel::detectCores() - 1L)
)
```

### Step 5 — Posterior probabilities and classification

```{r microc_probs, echo=TRUE, eval=FALSE}
microc_probs <- classify_hicpotts(microc_results, data = microc_data)

# Hard component assignment: the column with the highest posterior probability.
probs_mat <- as.matrix(microc_probs[, c("prob1", "prob2", "prob3")])
component <- max.col(probs_mat, ties.method = "first")
table(component)
```

### Step 6 — Visualize results

At fine resolution the heatmap is particularly useful for spotting local
clusters of high-probability interactions against the backdrop of raw counts.

```{r microc_plot, echo=TRUE, eval=FALSE}
# Dual-triangle heatmap: upper = prob2 (blue), lower = log1p(count) (grey)
microc_plot <- plot_upper_prob_lower_count(
    results       = microc_probs,
    bin1_col      = "start",
    bin2_col      = "start.j.",
    prob_col      = "prob2",
    count_col     = "interactions",
    chr_label     = "chr4",
    title         = "Micro-C significant interactions – BG3 chr4 (5 kb)",
    use_log_count = TRUE,
    symmetric_matrix = TRUE
)
print(microc_plot)
```

# Advanced Usage

`run_chain_betas()` is the public fitting interface. The examples in this
section use four independently seeded chains and the robust workflow so that
split-R-hat, effective sample size, component occupancy and allocation
diagnostics can be evaluated together.

There are two distinct prior workflows:

* With `use_data_priors = TRUE`, the robust workflow first runs a short
    multi-chain pilot, pools the pilot information into one empirical-Bayes
    prior, and freezes that prior for all production chains. A message such as
    `Running 4 shared-prior pilot chains for 5000 pilot iterations; production
    will then run up to 20000 requested iterations per chain.` is therefore
    expected only in this workflow. Pilot draws are not included in posterior
    summaries.
* With `use_data_priors = FALSE`, no empirical-Bayes pilot is needed. A
    supplied `user_fixed_priors` object is shared unchanged by every production
    chain. Consequently, the shared-pilot message is not printed. With
    `robust = TRUE` and `user_fixed_priors = NULL`, HiCPotts constructs fixed,
    covariate-scale-aware priors instead.

## Custom Priors

`user_fixed_priors` must be a named list containing `component1`, `component2`
and `component3`. Each component contains the prior mean and standard deviation
for the intercept and the four regression coefficients. The mapping is:

**`meany`, `sdy`**

intercept

**`meanx1`, `sdx1`**

genomic distance

**`meanx2`, `sdx2`**

GC content

**`meanx3`, `sdx3`**

transposable-element count

**`meanx4`, `sdx4`**

accessibility


All standard deviations must be finite and strictly positive. Values are on
the fitted coefficient scale (`log1p` covariates, without rewriting the
reported coefficients onto a standardized scale). Prior values should be
chosen from scientific knowledge or a justified external analysis; the broad
values below are only a template.

```{r fixed_priors, echo=TRUE, eval=FALSE}
# Extract the one-dataset matrices returned by process_data().
y_matrix <- processed[["y"]][[1L]]
x_matrices <- lapply(processed[["x_vars"]], function(x) x[[1L]])

base_component_prior <- list(
    meany = 0,
    meanx1 = 0, meanx2 = 0, meanx3 = 0, meanx4 = 0,
    sdy = 5,
    sdx1 = 1, sdx2 = 1, sdx3 = 1, sdx4 = 1
)

user_priors <- list(
    component1 = base_component_prior,
    component2 = base_component_prior,
    component3 = base_component_prior
)

# Components can have different priors. For example:
user_priors$component2$meany <- 1
user_priors$component2$sdy <- 2

fit_args <- list(
    N = nrow(y_matrix),
    x_vars = x_matrices,
    y = y_matrix,
    dist = "ZINB",
    gamma_start = 0.3,
    gamma_prior_shape = c(1, 1),
    iterations = 20000L,
    burnin = 5000L,
    n_chains = 4L,
    seeds = c(101L, 202L, 303L, 404L),
    robust = TRUE,
    initialization = "auto",
    use_data_priors = FALSE,
    user_fixed_priors = user_priors,
    theta_start = 0.05,
    size_start = rep(11, 3L),
    mcse_stop = FALSE,
    mc_cores = 4L,
    verbose = TRUE,
    progress_interval = 1000L
)

fit <- do.call(run_chain_betas, fit_args)
```

`gamma_start` is the initial Potts parameter value. It is not the prior on
gamma; `gamma_prior_shape = c(a, b)` supplies the two shape parameters of its
Beta prior. Setting `mcse_stop = FALSE` forces every production chain to run
the requested 20,000 iterations. When `mcse_stop = TRUE`, `iterations` is a
maximum and a chain may stop after satisfying the configured MCSE rule.

## Independent Chains and Parallel Processing

`n_chains` controls how many independently seeded chains are fitted for each
dataset. `mc_cores` controls how many of those chains can run concurrently; it
does not change the number of chains or the posterior target. Four or more
chains are recommended when convergence will be assessed. Reproducible seeds
should be supplied explicitly.

```{r parallel_chains, echo=TRUE, eval=FALSE}
fit <- run_chain_betas(
    N = nrow(y_matrix),
    x_vars = x_matrices,
    y = y_matrix,
    dist = "ZIP",
    gamma_start = 0.3,
    theta_start = 0.05,
    iterations = 20000L,
    burnin = 5000L,
    n_chains = 4L,
    seeds = c(101L, 202L, 303L, 404L),
    robust = TRUE,
    initialization = "auto",
    use_data_priors = TRUE,
    mc_cores = 4L
)
```

For one dataset, `y` can be a matrix and each `x_vars` entry can be a matrix.
For several datasets, `y` must be a list of count matrices and every named
covariate entry must be a list with one matrix per dataset. All matrices must
be square, finite, and use the same `N`. The distribution is common to all
datasets in one call. A supplied fixed-prior object is also reused for every
dataset; fit datasets separately if they require different fixed priors.

```{r multiple_matrices, echo=TRUE, eval=FALSE}
y_many <- list(sample_A = y_A, sample_B = y_B)
x_many <- list(
    distance = list(distance_A, distance_B),
    GC = list(GC_A, GC_B),
    TES = list(TES_A, TES_B),
    ACC = list(ACC_A, ACC_B)
)

fits_many <- run_chain_betas(
    N = N,
    x_vars = x_many,
    y = y_many,
    dist = "NB",
    gamma_start = 0.3,
    size_start = rep(11, 3L),
    iterations = 20000L,
    burnin = 5000L,
    n_chains = 4L,
    seeds = c(101L, 202L, 303L, 404L),
    robust = TRUE,
    use_data_priors = TRUE,
    mc_cores = 4L
)
```

## Other Distributions

The count family controls how sampling variation, over-dispersion and excess
zeros are modelled. It does not change the three biological component labels.

**`"Poisson"`**

Default: `equidispersed counts`. none

**`"NB"`**

Default: `over-dispersed counts`. `size_start`, length three

**`"ZIP"`**

Default: `Poisson counts with excess zeros`. `theta_start`

**`"ZINB"`**

Default: `over-dispersed counts with excess zeros`. both


Use posterior-predictive checks and convergence diagnostics to compare
families. Do not choose a family solely because it produces more interactions
in one latent class.

```{r all_distributions, echo=TRUE, eval=FALSE}
fit_one_family <- function(family) {
    args <- list(
        N = nrow(y_matrix),
        x_vars = x_matrices,
        y = y_matrix,
        dist = family,
        gamma_start = 0.3,
        iterations = 20000L,
        burnin = 5000L,
        n_chains = 4L,
        seeds = c(101L, 202L, 303L, 404L),
        robust = TRUE,
        use_data_priors = FALSE,
        user_fixed_priors = user_priors,
        mcse_stop = FALSE,
        mc_cores = 4L
    )
    if (family %in% c("ZIP", "ZINB")) args$theta_start <- 0.05
    if (family %in% c("NB", "ZINB")) args$size_start <- rep(11, 3L)
    do.call(run_chain_betas, args)
}

fits_by_family <- setNames(
    lapply(c("Poisson", "NB", "ZIP", "ZINB"), fit_one_family),
    c("Poisson", "NB", "ZIP", "ZINB")
)
```

## Sampler Accuracy and Runtime Controls

The ABC update for the Potts parameter is usually the dominant cost on large
lattices. The following controls are exposed by `run_chain_betas()`:

* `abc_potts_sweeps = 0L` selects the lattice-dependent default. Supplying a
    smaller positive value is faster but may under-equilibrate auxiliary Potts
    fields and bias spatial coupling upward.
* `abc_sim_reps` is the number of auxiliary fields averaged per gamma proposal.
    More replicates reduce decision noise and increase cost approximately
    linearly.
* `gamma_update_interval` is the number of MCMC iterations between gamma
    proposals. Larger values are cheaper but yield fewer gamma transitions.
* `epsilon = NULL` calibrates one ABC bandwidth and shares it across production
    chains. A fixed positive value is appropriate only when it has been
    justified
    for the lattice and summary statistic.
* `mcse_min_iterations`, `mcse_check_interval` and
    `mcse_relative_threshold` configure optional early stopping. Use
    `mcse_stop = FALSE` when an exact iteration count is required.

These parameters trade computation for Monte Carlo accuracy. Any non-default
choice should be recorded and followed by checks of gamma movement, ESS,
split-R-hat and allocation uncertainty.

## Diagnostics and Parameter Summaries

The robust result records both selected chains and reliability flags. Run the
official classification before attaching probability diagnostics, then inspect
the gates before reporting parameters.

```{r advanced_diagnostics, echo=TRUE, eval=FALSE}
classification <- classify_hicpotts(
    fit,
    data = data,
    min_draws = 100L,
    reflect = "auto"
)

diagnostics <- diagnose_hicpotts_fit(
    fit,
    prob_result = classification
)

allocation <- allocation_diagnostics(fit)

# Apply the configured parameter-reporting criteria.
parameter_summary <- summarise_hicpotts_parameters(
    fit,
    x_vars = x_matrices,
    require_reliable = TRUE
)

# Return the table together with row-level diagnostic indicators.
diagnostic_summary <- summarise_hicpotts_parameters(
    fit,
    x_vars = x_matrices,
    require_reliable = FALSE
)

ppc <- posterior_predictive_hicpotts(
    fit,
    x_vars = x_matrices,
    y = y_matrix,
    dist = "ZINB",
    n_rep = 100L,
    seed = 1L
)
```

The diagnostic table records each criterion separately, so the result can be
read alongside the specified thresholds, retained chain length and trace
plots. Report these settings and the corresponding ESS/R-hat values with
parameter estimates.

## Symmetric and Ordered Matrices

For mirrored or triangular Hi-C input, use `symmetric_matrix = TRUE`. The plot
then uses one representative of each unordered bin pair: the visually upper
triangle shows the requested posterior probability and the visually lower
triangle, including the diagonal, shows observed counts. This intentionally
produces a dual-triangle display rather than mirroring the same quantity into
both halves.

Use `symmetric_matrix = FALSE` only when `(bin1, bin2)` and `(bin2, bin1)` are
meaningfully different ordered observations. In that mode no mirror-based
deduplication is performed.

```{r advanced_symmetric_plot, echo=TRUE, eval=FALSE}
plot_upper_prob_lower_count(
    results = classification,
    bin1_col = "start",
    bin2_col = "start.j.",
    prob_col = "prob2",
    count_col = "interactions",
    symmetric_matrix = TRUE,
    use_log_count = TRUE,
    chr_label = "chr4",
    title = "Signal probability and observed counts"
)
```

# Technical Notes

Performance: The allocation, Potts-neighbour and MCMC kernels are implemented
internally in C++ using Rcpp and RcppArmadillo. Users call `run_chain_betas()`
and choose `n_chains` and `mc_cores`; native kernels and validation helpers are
intentionally not exported.

Visualization: `plot_upper_prob_lower_count()` creates a dual-triangle Hi-C
heatmap from the data frame returned by `classify_hicpotts()`. It requires the
`ggplot2` (>= 3.5.0) and `ggnewscale` packages, both of which are listed as
dependencies of HiCPotts. The `prob_col` argument selects which probability
column to display in the upper triangle (default `"prob2"`); `use_log_count`
controls whether raw or `log1p`-transformed counts appear in the lower triangle.
The `bin1_col` and `bin2_col` arguments should name the columns holding the
genomic start positions of the two interacting bins — typically `"start"` and
`"start.j."` in `get_data()` output.

# Conclusion

The HiCPotts package offers a powerful tool for Bayesian analysis of Hi-C data,
integrating spatial dependencies, information from sources of bias associated
with Hi-C data, and flexible mixture models. This page covered the core
workflow, but the package’s functions can be customized for specific research
needs, such as different genomic regions or distribution assumptions.

Users trying to bring in extra covariates; beyond genomic distance, GC content,
ACC-score and TEs, can do so by classifying each new factor according to how it
relates to those four (for example, as a distance-like term, a
sequence-composition term, or an interaction‐score term) and then inputting it
into the same model framework. Also, the package currently supports only
intrachromosomal analyses. In the future, we plan to extend to interchromosomal
contacts.


For further details, consult the package documentation (`?HiCPotts`) or contact
the package maintainers. We hope HiCPotts facilitates your genomic research!



```{r session_info, echo = FALSE}
sessionInfo()
```
